Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution

### 2021-08-04 Paper/Submission_new/Main_Document-Latex.tex
04.08: Revisions first round completed

1. Use of invertebrate traits rather than species composition may facilitate large-scale comparisons of community structure and responses to disturbance in freshwater ecology because the same traits can potentially occur everywhere.  In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially at large scales. However, a number of data-related issues complicate the use of invertebrate traits for ecological studies. For example, standardised definitions for freshwater invertebrate traits are lacking, which impedes comparisons across regions. Moreover, it is uncertain how harmonising varying trait definitions among databases might influence identification of trait-environment relationships. In addition, taxonomic resolution between taxonomic composition data and trait data can differ, making aggregation of traits necessary. At present, it is unknown how different trait aggregation approaches compare with expert-assigned trait affinities. In this study we aimed to identify discrepancies in trait definitions across freshwater invertebrate databases and to compare trait aggregation methods to identify how phylogenetic structure and trait variability influence results from different aggregation methods.

@@

3. We found that among trait databases, trait definitions often differed because varying numbers of traits were used to describe groups of related traits (e.g., respiration traits) and the focus of descriptions for groups of related traits also varied (e.g., for feeding mode some databases focused on the food source, whereas others focused on mouthpart morphology). The coding to describe traits (binary, fuzzy) also varied among databases. Our comparison of different aggregation methods showed that family-level aggregated and expert-assigned traits were similar, especially when aggregation was based on the median, and that trait coding method, uneven phylogenetic structure, and high trait variability can influence aggregated trait affinities. We further showed that harmonised and aggregated data performed similarly to not aggregated and harmonised data from the case study in identifying trait–environment relationships.

4. By identifying discrepancies in trait definitions we hope to motivate the development of standardised terminology for invertebrate traits. Our results also illustrate the usefulness of harmonised datasets for ecological study and provide guidance for the circumstances under which the choice of trait aggregation method is important.
\\
\keywords{trait definitions, trait aggregation, data synthesis, trait-environment relationships, large-scale comparisons}

@@

Invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors such as climate change or salinisation (\cite{bhowmik_large_2015, poff_developing_2010, szocs_effects_2014}), for predicting changes in ecosystem functioning and trophic dynamics (\cite{vos_taxonomic_2017, gutierrez-canovasPopulationsHighvaluePredators2021}), or for developing indices for biomonitoring (\cite{doledecInvertebrateTraitsBiomonitoring2008a, beketov_spear_2009}), but see also \citet{hamiltonLimitationsTraitbasedApproaches2020} for a critical perspective. In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases, which typically include data for a single region or continent (\cite{kefford_integrated_2020, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate, gayraudInvertebrateTraitsBiomonitoring2003, statznerConservationTaxonomicBiological2007}). The availability of invertebrate trait data from different continents enables comparisons of regional trait variation and allows for testing the consistency of trait structure–environmental factor relationships across both small and large spatial extents. 
% Traits provide adaptations to the environmental variability of a habitat which may explain species distributions and early work focused on the link between habitat utilization and traits which lead to one of the first collections of trait information on aquatic invertebrates 
% See theoretical habitat templets (Polatera & Tachet 1994)

To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterised trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses that synthesise invertebrate trait information from more than two different continents are rare, but see \citet{brown_functional_2018} and \citet{statzner_reproductive_1997}.

The variability and inconsistencies of information in freshwater invertebrate trait databases, besides the diversity of taxa across regions, is likely a major reason for the lack of comparisons across continents. Several challenges currently prevent the use of invertebrate trait data from different established databases in ecological studies. One challenge is inconsistency in terminology. In this study, we follow the terminology proposed by \citet{schmera_proposed_2015}, in which a grouping feature (e.g. feeding mode) refers to a set of related traits (e.g. predator, shredder, etc.) that together define a general property (e.g. respiration) that is shared among individuals or species. Thus, to use consistent terminology in this study, we use the term grouping feature in place of the term trait, and the term trait instead of trait state, modality or trait category.

@@

A third challenge to comparing across invertebrate trait databases is inconsistent coding of trait data (\cite{culp_incorporating_2011}). Traits of individual freshwater invertebrates can be difficult to quantify because, unlike plant traits, they are often difficult to measure directly. For example, to describe feeding habits requires that we understand mouthpart morphology, feeding behaviour, and what is being eaten (\cite{moog_comprehensive_nodate}). Some traits can be described in categorical terms, which ignores uncertainty in how traits are expressed in any particular organism (e.g., adult terrestrial stage, presence of gills). Other traits are better suited to being described as continuous (e.g., tolerance of pollution, body size). One approach for dealing with uncertainty is the use of fuzzy coding, where traits are assigned probabilistic values (\cite{chevenet_francois_fuzzy_1994}). Fuzzy codes, which are usually converted to proportions, are used to account for plasticity in traits, variability in traits within taxonomic groups above the species level, and incomplete knowledge. The values derived from the different coding approaches represent affinities (aka affinity scores) that indicate how strongly a taxon expresses a particular trait. The use of different types of coding for trait affinities in different databases limits comparisons. For example, in the study from \citet{brown_functional_2018} the authors needed to reclassify trait affinities because the individual databases employed different coding approaches (i.e., European and New Zealand databases used fuzzy coding whereas the North American database used binary coding).

@@

We assessed how six databases differed in how they assigned and described traits of freshwater invertebrates. These databases (from Europe, North America, Australia, and New Zealand) differed in how grouping features were defined and the taxonomic resolution used to assign traits. For Europe we obtained trait information from the freshwaterecology.info database (\cite{schmidt-kloiber_www.freshwaterecology.info_2015}), which assigns traits at the species level, and the Tachet database (\cite{ussegliopolatera_biological_2000}), which assigns traits at species, genus, and family levels. For North America we primarily obtained trait information from \citet{twardochleb_freshwater_nodate} (hereafter referred to as the CONUS database) and where taxa were missing from that database, we obtained data from \citet{vieira_database_nodate} (hereafter referred to as the Vieira database). Both databases contain taxa at species, genus and family levels. Data on body form for European and North American taxa were based on expert knowledge (\cite{polatera_personal_information_2020}). For Australia and New Zealand, we used trait databases from \citet{kefford_integrated_2020} (hereafter referred to as the Australian database) and \citet{Philips_and_Smith_NZ_DB_2018} (hereafter referred to as the New Zealand database). These databases assigned traits at species, genus, family, and coarser taxonomic levels.

We established four datasets, one for each of the four geographic regions, by harmonising the traits of seven grouping features from the trait databases. We selected grouping feautres that are commonly applied in trait-based ecological studies and describe different aspects of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode), and reproduction (Oviposition). A criterion for trait selection was that the traits occurred in all databases, to enable us to show trait definitions differences and the influence of grouping-feature harmonisation on trait-environment relationships. Thus, we only used biological traits and omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database. Also, traits from the selected grouping features may consistently respond to certain environmental stressors across large scales and thus be suitable for intercontinental comparisons of trait-environment relationships. For example, traits like body size small, multiple life cycles per year, or swimming are related to resilience and resistance to flow disturbances or retreating glacier cover and may consistently respond to certain environmental stressors across regions (\cite{bonada_taxonomic_2007, brown_functional_2018}).

Grouping features were classified differently across the databases, so we harmonised 7 grouping features that collectively included 26 traits (Table \ref{tab:traits_harmonisation}). To do so we combined similar traits into one trait (e.g. crawlers and sprawlers into crawlers) by assigning for each taxon for the new combined trait the highest affinity score of those originally assigned. Next, we selected taxa for inclusion in our datasets and normalised the trait affinity scores for each taxon. We first omitted all taxa with a lower taxonomic precision than family level and consolidated duplicate taxa that appeared within a database, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We then used either fuzzy coding or, when unavailable, binary coding to create the final harmonised datasets. Fuzzy codes are reported with different ranges in the trait databases (e.g., freshwaterecology.info: 0–10, Tachet database: 0–3 or 0–5). We normalised these to a range between 0 and 1 and converted trait affinities to proportions. In cases where binary coding was needed, we converted categorical and continuous traits into binary traits with values of 1 and 0, which, therefore, had the same range as fuzzy coded traits. For example, in the  freshwaterecology.info database, voltinism traits are associated with different faunistic regions, and we substituted a value of 1 for entries such as “arctic” or “boreal”. We assumed that values of 1 and 0 for binary traits corresponded to the highest and lowest affinity for a particular trait.

@@

Further, we examined the influence of phylogeny and trait variability on the outcomes of the different trait aggregation methods by simulating families with different phylogenetic structures and different levels of trait variability. The simulation showed that both phylogenetic structure and trait variability can affect aggregated trait affinities, but mostly only to a small degree. As expected, the range of trait affinities increased with higher trait variability for all trait aggregation methods. Therefore, we provide details of the simulation and the results in the supplementary materials.

@@

We re-analysed the \citet{szocs_effects_2014} data by performing a RDA with a mixture of harmonised grouping features from the dataset for Europe and non-harmonised grouping features, and with traits that we aggregated to family level with the five aggregation methods described above. We used all 21 grouping features from the original study, but replaced 6 with harmonised grouping features from the established European trait dataset, namely size, feeding mode, locomotion, oviposition, respiration, and voltinism.
We compared the results of the original study’s RDA with the results of the RDA from the harmonised grouping features. Specifically, trait composition, expressed as community weighted mean (CWM) traits, was ordinated along an electrical conductivity gradient. We compared the species scores obtained from the RDA (i.e., the coordinates of the tips of the vectors representing the CWM traits in the bi- or triplots). Following the original study, we used the Mahalanobis distance measure and identified traits associated with higher or lower salinity based on their distance to the ordination axis median. We considered CWM traits with a Mahalanobis distance greater than the 97.5 \%-quantile of the Chi-square distribution (5.02) to have responded to either lower or higher salinity. Additionally, to test for the effect of different aggregation methods on inferences about CWM trait–salinisation relationships, we repeated the RDA five times, each with family-level aggregated trait values assigned to corresponding taxa from \citet{szocs_effects_2014}. We used the RDA approach for consistency with \citet{szocs_effects_2014} but are aware of the criticism that regressing CWM traits against environmental gradients has received (but see \cite{peres-netoLinkingTraitVariation2017, millerFunctionalTraitsCommunity2019}).
% because CWM trait values are treated as independent while they in fact are non-independent which can lead to inflated Type 1 errors

Finally, we assessed how functional diversity (FD) changed when harmonised and aggregated trait data are used. Harmonisation of grouping features reduces the number of traits per grouping feature, e.g. for the freshwaterecology.info database ten feeding mode traits were combined to six traits, and thus likely influences FD metrics. We calculated functional richness (FRic), functional evenness (FEve), and functional divergence (FDiv) following \citet{villegerNewMultidimensionalFunctional2008a} with the traits from \citet{szocs_effects_2014}, harmonised grouping features from the European trait dataset, and with family-level aggregated trait values. For FEve and FDiv invertebrate abundance data were used from \citet{szocs_effects_2014}. To account for temporal autocorrelation (sampling was done over a time period of three years) we regressed each FD metric with year and extracted the residuals. The residuals of the original study were compared with the residuals of the harmonised and aggregated datasets through regression and evaluation of the coefficient of determination ($R^2$).

@@

Definitions of grouping features and traits varied in their level of detail in the six trait databases. The freshwaterecology.info, Tachet, and CONUS databases provided more detailed descriptions of trait information than the Vieira and New Zealand databases provided. The Australian trait database (\cite{kefford_integrated_2020}) was a collection of seven specific trait datasets; thus, grouping features occurred multiple times with varying differentiation into traits. Depending on the dataset, trait information was described with more or less detail.

@@

The percentage of differing cases of aggregated versus assigned trait affinities varied by method between 16.2 and 22.9 \% for the Australian dataset and between 15.3 and 47 \% for the North American dataset (Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). In general, median aggregation methods yielded fewer cases with differences compared with mean aggregation methods. However, median aggregation methods produced, on average, greater absolute differences for both datasets. Standard deviations of absolute differences were similar for all tested aggregation methods. Maximum differences of 1 (i.e., the maximum possible difference) occurred for all investigated grouping features in both datasets (Figure \ref{fig:diff_aggr_traits_combined}).

@@

RDA based on harmonised traits that were aggregated to family level also produced results similar to the original analysis but with fewer traits distinguishing upstream and downstream sites (Figures \ref{fig:violin_plot_species_sc} and \ref{fig:boxplot_scores_on_constrained_axis}). The \textit{direct\_agg\textsubscript{mean}}, \textit{direct\_agg\textsubscript{median}}, and \textit{weighted\_agg} characterised the downstream sites with the same traits as the original analysis except the trait shredder. As for the RDA with harmonised but not aggregated data, RDA with these three aggregation methods characterised upstream sites by the traits univoltinism, short life-cycle duration, and aquatic eggs. The \textit{stepwise\_agg\textsubscript{mean}} and \textit{stepwise\_agg\textsubscript{median}} methods produced the same results except that none of the life-cycle traits characterised upstream or downstream sites. Thus, compared with the original study’s findings, aggregating by the direct and weighted methods yielded less change than the stepwise methods in interpretation of the RDA results. To demonstrate a different approach to detecting relationships between traits and conductivity we used fourth corner analysis with the original, harmonised, and aggregated trait datasets (methods and results in  ).

@@

The calculation of FD metrics for the original, harmonised (not aggregated), and aggregated datasets suggested that harmonising trait data preserves FRic and FDiv values, while aggregated trait data only preserved FRic values. The FRic and FDiv values calculated from harmonised trait data correlated highly to FRic and FDiv values from the original dataset ($R^2 = 0.96$ and $0.86$, respectively). FRic and FDiv calculated for aggregated trait data showed moderate to good correlations to the FRic and FDiv values calculated with the original dataset (FRic: $0.78 <= R^2 <= 0.84$ , FDiv: $0.42 <= R^2 <= 0.62$, Figure \ref{fig:FRic} and \ref{fig:FDiv}). By contrast, FEve values calculated with harmonised and aggregated trait data were unrelated to the FEve values calculated with the original dataset (all $R^2 < 0.1$), except for \textit{stepwise\_agg \textsubscript{median}}, which showed a moderate correlation to FEve values calculated from the original dataset ($R^2  = 0.48$, Figure \ref{fig:FEve}).

@@

FD analysis is an important trait-based application to quantify the variability of traits that influence ecosystem functions. We found through correlation that harmonised trait data may be used to calculate FRic and FDiv values, and that aggregated trait data can at least be used to calculate FRic values. This does not apply to FEve, as the use of harmonised and aggregated trait data yielded completely different FEve values than with the trait data from \citet{szocs_effects_2014}. FRic represents the trait space spanned by the community (also termed the convex hull volume) and is only dependent on the trait data used. Our results show that harmonisation preserves the trait space the best, i.e. the most extreme trait values of the original study, and that aggregation only slightly changes the trait space. It follows that FDiv, which describes how abundances are distributed within the volume of the trait space, is best preserved by harmonised trait data and to a lesser extent with aggregated trait data. FEve describes how regular the trait space is filled by taxa. We found that harmonised and aggregated trait change the evenness of the abundance distribution in the trait space, which is likely because during the calculation of FEve only the nearest neighbours in the trait space are considered. This leads to the undesirable property that even small community changes result in large changes in the FEve index, a behaviour that has been shown by \citet{botta-dukatTestingAbilityFunctional2016} through simulation.

@@

Although comprehensive freshwater invertebrate trait databases have been developed for several regions in the past, data synthesis presents challenges because of discrepancies in trait definitions, coding of trait affinities, and taxonomic resolution. By providing an overview of definition discrepancies, we have set a starting point for the development of standardised trait terminology through which invertebrate trait databases can be linked. Agreement on standard terminology and the subsequent development of ontologies are the next steps to facilitate trait-based analysis at large geographical scales. As our analysis showed, some grouping features might need to be re-classified to fit into such a standardised terminology. Although we showed that trait affinities from fuzzy and binary coding can be used together, we also recommend that a uniform coding of traits should be addressed during trait standardisation.

@@

\section*{Author contribution}
R.B.S. and S.K. designed the study; B.J.K. and L.M. provided trait data for Australia; P.U-P. provided body form trait data for European and North American invertebrates; C.P.H., N.L.P., and L.T. provided trait data for North America; C.D.M. provided trait data for New Zealand;
B.J.K., A.S-K., C.D.M, W.G., N.L.P., L.M., and C.P.H. provided in depth information on trait data and discussed the procedure of analysis; S.K. conducted data processing and the statistical analyses.; S.K. wrote the manuscript; B.J.K, A-S.K., C.D.M, P.U-P., C.P.H., and R.B.S. discussed the results and provided suggestions on the manuscript.

>>>>

1. Use of invertebrate traits rather than species composition may facilitate large-scale comparisons of community structure and responses to disturbance in freshwater ecology because the same traits can potentially occur everywhere.  In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially at large scales. However, a number of data-related issues complicate the use of invertebrate traits for ecological studies. For example, standardised definitions for freshwater invertebrate traits are lacking, which impedes comparisons across regions. Moreover, it is uncertain how harmonising varying trait definitions among databases might influence identification of trait-environment relationships. In addition, taxonomic resolution between taxonomic composition data and trait data can differ, making aggregation of traits necessary. However, at present, it is unknown how different trait aggregation approaches compare with expert-assigned trait affinities.
% In this study we aimed to identify discrepancies in trait definitions across freshwater invertebrate databases and to compare trait aggregation methods to identify how phylogenetic structure and trait variability influence results from different aggregation methods.

@@

3. We found that among trait databases, trait definitions often differed because varying numbers of traits were used to describe groups of related traits (e.g., respiration traits) and the focus of descriptions for groups of related traits also varied (e.g., for feeding mode some databases focused on the food source, whereas others focused on mouthpart morphology). The coding to describe traits (binary, fuzzy) also varied among databases. Our comparison of different aggregation methods showed that family-level aggregated and expert-assigned traits were similar, especially when aggregation was based on the median.
% ,and that trait coding method, uneven phylogenetic structure, and high trait variability can influence aggregated trait affinities. 
We further showed that harmonised and aggregated data performed similarly to non-aggregated and harmonised data from the case study in identifying trait–environment relationships, but were only partially able to yield similar values for functional diversity metrics in the case study.

4. By identifying discrepancies in trait definitions we hope to motivate the development of standardised definitions for invertebrate traits. Our results also illustrate the usefulness of harmonised datasets for ecological study and provide guidance for the circumstances under which the choice of trait aggregation method is important.
\\
\keywords{trait definitions, trait aggregation, data synthesis, trait-environment relationships, large-scale comparisons}

@@

Invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors such as climate change or salinisation (\cite{bhowmik_large_2015, contiTraitbasedApproachAssess2014, poff_developing_2010, szocs_effects_2014}), for predicting changes in ecosystem functioning and trophic dynamics (\cite{gutierrez-canovasPopulationsHighvaluePredators2021, vos_taxonomic_2017}), for developing indices or diagnostic tools for biomonitoring (\cite{beketov_spear_2009, doledecInvertebrateTraitsBiomonitoring2008a, mondyUsingConditionalTree2013}), but see also \citet{hamiltonLimitationsTraitbasedApproaches2020} for a critical perspective. In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases, which typically include data for a single region or continent (\cite{gayraudInvertebrateTraitsBiomonitoring2003, keffordIntegratedDatabaseStream2020a, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015,
statznerConservationTaxonomicBiological2007, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate}). The availability of invertebrate trait data from different continents enables comparisons of regional trait variation and allows for testing the consistency of trait structure–environmental factor relationships across both small and large spatial extents. In fact, the first studies on the link between habitat utilization and traits in the Upper Rhône river initiated the compilation of aquatic invertebrate trait information (\cite{bournaudApproachSynthesisQualitative1992,usseglio-polateraTheoreticalHabitatTemplets1994}). At large scales, consistent responses of traits to environmental stressors have been reported, e.g. traits like small body size, multiple life cycles per year, or swimming have been related to resilience and resistance to flow disturbances or retreating glacier cover (\cite{bonada_taxonomic_2007, brown_functional_2018}). To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterised trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses that synthesise invertebrate trait information from more than two different continents are rare, but see \citet{brown_functional_2018} and \citet{statzner_reproductive_1997}.

The variability and inconsistencies of information in freshwater invertebrate trait databases, besides the diversity of taxa across regions, is likely a major reason for the lack of comparisons across continents. Several challenges currently prevent the use of invertebrate trait data from different established databases in ecological studies. One challenge is inconsistency in terminology. In this study, we follow the terminology proposed by \citet{schmera_proposed_2015}, in which a grouping feature (e.g. feeding mode) refers to a set of related traits (e.g. predator, shredder, etc.) that together define a general property that is shared among individuals or species. Thus, to use consistent terminology in this study, we use the term grouping feature in place of the term trait, and the term trait instead of trait state, modality or trait category.

@@

A third challenge to comparing across invertebrate trait databases is inconsistent coding of trait data (\cite{culp_incorporating_2011}). Traits of individual freshwater invertebrates can be difficult to quantify because, unlike plant traits, they are often difficult to measure directly. For example, to describe feeding habits requires that we understand mouthpart morphology, feeding behaviour, and what is being eaten (\cite{moogFaunaAquaticaAustriaca2017}). Some traits can be described in categorical terms, which ignores uncertainty in how traits are expressed in any particular organism (e.g., adult terrestrial stage, presence of gills). Other traits are better suited to being described as continuous (e.g., tolerance of pollution, body size). One approach for dealing with uncertainty is the use of fuzzy coding, where traits are assigned probabilistic values (\cite{chevenet_francois_fuzzy_1994}). Fuzzy codes, which are usually converted to proportions, are used to account for plasticity in traits, variability in traits within taxonomic groups above the species level, and incomplete knowledge. The values derived from the different coding approaches represent affinities (aka affinity scores) that indicate how strongly a taxon expresses a particular trait. The use of different types of coding for trait affinities in different databases limits comparisons. For example, in the study from \citet{brown_functional_2018} the authors needed to reclassify trait affinities because the individual databases employed different coding approaches (i.e., European and New Zealand databases used fuzzy coding whereas the North American database used binary coding).

@@

We assessed how six databases differed in how they assigned and described traits of freshwater invertebrates. These databases (from Europe, North America, Australia, and New Zealand) differed in how grouping features were defined and the taxonomic resolution used to assign traits. For Europe we obtained trait information from the freshwaterecology.info database (\cite{schmidt-kloiber_www.freshwaterecology.info_2015}), which assigns traits at the species level, and the Tachet database (\cite{ussegliopolatera_biological_2000}), which assigns traits at species, genus, and family levels. For North America we primarily obtained trait information from \citet{twardochleb_freshwater_nodate} (hereafter referred to as the CONUS database) and where taxa were missing from that database, we obtained data from \citet{vieira_database_nodate} (hereafter referred to as the Vieira database). Both databases contain taxa at species, genus and family levels. Data on body form for European and North American taxa were based on expert knowledge (\cite{polatera_personal_information_2020}). For Australia and New Zealand, we used trait databases from \citet{keffordIntegratedDatabaseStream2020a} (hereafter referred to as the Australian database) and \citet{Philips_and_Smith_NZ_DB_2018} (hereafter referred to as the New Zealand database). These databases assigned traits at species, genus, family, and coarser taxonomic levels.

We established four datasets, one for each of the four geographic regions, by harmonising the traits of seven grouping features from the trait databases. We selected grouping features that are commonly applied in trait-based ecological studies and describe different aspects of the biology of a species: life history (Voltinism), morphology (Body form, Size), physiology (Feeding mode, Respiration), mobility (Locomotion), and reproduction (Oviposition). A criterion for trait selection was that the traits occurred in all databases, to allow for an analysis of potential differences in trait definitions and of the influence of grouping-feature harmonisation on trait-environment relationships. Thus, we only used biological traits and omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database.

Grouping features were classified differently across the databases, so we harmonised 7 grouping features that collectively included 26 traits (Table \ref{tab:traits_harmonisation}). To do so we normalised trait affinity scores and combined similar traits into one trait (e.g. crawlers and sprawlers into crawlers) by summing for each taxon for the new combined trait the affinity scores of those originally assigned. If similar traits from multiple databases were combined (e.g. in the Australian trait database) we assigned the highest affinity score. Next, we selected taxa for inclusion in our datasets and normalised the trait affinity scores for each taxon again. We first omitted all taxa with a lower taxonomic precision than family level and consolidated duplicate taxa that appeared within a database, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We then used either fuzzy coding or, when unavailable, binary coding to create the final harmonised datasets. Fuzzy codes are reported with different ranges in the trait databases (e.g., freshwaterecology.info: 0–10, Tachet database: 0–3 or 0–5). We normalised these to a range between 0 and 1 and converted trait affinities to proportions. In cases where binary coding was needed, we converted categorical and continuous traits into binary traits with values of 1 and 0, which, therefore, had the same range as fuzzy coded traits. For example, in the  freshwaterecology.info database, voltinism traits are associated with different faunistic regions, and we substituted a value of 1 for entries such as “arctic” or “boreal”. We assumed that values of 1 and 0 for binary traits corresponded to the highest and lowest affinity for a particular trait.

@@

Further, we examined the influence of phylogeny and trait variability on the outcomes of the different trait aggregation methods by simulating families with different phylogenetic structures and different levels of trait variability. The simulation showed that both phylogenetic structure and trait variability can affect aggregated trait affinities, but mostly only to a small degree. As expected, the range of trait affinities increased with higher trait variability for all trait aggregation methods. We provide details of the simulation and the results in the Supplementary Materials.

@@

We re-analysed the \citet{szocs_effects_2014} data by performing a RDA with a mixture of harmonised grouping features from the dataset for Europe and non-harmonised grouping features, and with traits that we aggregated to family level with the five aggregation methods described above. We used all 21 grouping features from the original study, but replaced 6 with harmonised grouping features from the established European trait dataset, namely size, feeding mode, locomotion, oviposition, respiration, and voltinism.
We compared the results of the original study’s RDA with the results of the RDA from the harmonised grouping features. Specifically, trait composition, expressed as community weighted mean (CWM) traits, was ordinated along an electrical conductivity gradient. We compared the species scores obtained from the RDA (i.e., the coordinates of the tips of the vectors representing the CWM traits in the bi- or triplots). Following the original study, we used the Mahalanobis distance measure and identified traits associated with higher or lower salinity based on their distance to the ordination axis median. We considered CWM traits with a Mahalanobis distance greater than the 97.5 \%-quantile of the Chi-square distribution (5.02) to have responded to either lower or higher salinity. Additionally, to test for the effect of different aggregation methods on inferences about CWM trait–salinisation relationships, we repeated the RDA five times, each with family-level aggregated trait values assigned to corresponding taxa from \citet{szocs_effects_2014}. We used the RDA approach for consistency with \citet{szocs_effects_2014}, though it is under debate (see \citet{peres-netoLinkingTraitVariation2017, millerFunctionalTraitsCommunity2019}).
% because CWM trait values are treated as independent while they in fact are non-independent which can lead to inflated Type 1 errors

Finally, we assessed how functional diversity (FD) changed when harmonised and aggregated trait data are used. Harmonisation of grouping features reduces the number of traits per grouping feature and thus likely influences FD metrics, e.g. for the freshwaterecology.info database ten feeding mode traits were combined to six traits. We calculated functional richness (FRic), functional evenness (FEve), and functional divergence (FDiv) following \citet{villegerNewMultidimensionalFunctional2008a} with the traits from \citet{szocs_effects_2014}, harmonised grouping features from the European trait dataset, and with family-level aggregated trait values. For FEve and FDiv invertebrate abundance data were used from \citet{szocs_effects_2014}. To account for temporal autocorrelation (sampling was done over a time period of three years), we partialled out the effect of year from all datasets. The datasets were compared through regression with the coefficient of determination ($R^2$).

@@

Definitions of grouping features and traits varied in their level of detail in the six trait databases. The freshwaterecology.info, Tachet, and CONUS databases provided more detailed descriptions of trait information than the Vieira and New Zealand databases provided. The Australian trait database (\cite{keffordIntegratedDatabaseStream2020a}) was a collection of seven specific trait datasets; thus, grouping features occurred multiple times with varying differentiation into traits. Depending on the dataset, trait information was described with more or less detail.

@@

The percentage of differing cases of aggregated versus assigned trait affinities varied by method between 21.9 and 27.6 \% for the Australian dataset and between 15.5 and 44.8 \% for the North American dataset (Table \ref{tab:summary_stat_aggr_vs_fam_assigned}). In general, median aggregation methods yielded fewer cases with differences compared with mean aggregation methods. However, median aggregation methods produced, on average, greater absolute differences for both datasets. Standard deviations of absolute differences were similar for all tested aggregation methods. Maximum differences of 1 (i.e., the maximum possible difference) occurred for all investigated grouping features in both datasets (Figure \ref{fig:diff_aggr_traits_combined}).

@@

RDA based on harmonised traits that were aggregated to family level also produced results similar to the original analysis but with fewer traits distinguishing upstream and downstream sites (Figures \ref{fig:violin_plot_species_sc} and \ref{fig:boxplot_scores_on_constrained_axis}). The \textit{direct\_agg\textsubscript{mean}}, \textit{direct\_agg\textsubscript{median}}, and \textit{weighted\_agg} characterised the downstream sites with the same traits as the original analysis except the trait shredder. As for the RDA with harmonised but not aggregated data, RDA with these three aggregation methods characterised upstream sites by the traits univoltinism, short life-cycle duration, and aquatic eggs. The \textit{stepwise\_agg\textsubscript{mean}} and \textit{stepwise\_agg\textsubscript{median}} methods produced the same results except that none of the life-cycle traits characterised upstream or downstream sites. Thus, compared with the original study’s findings, aggregating by the direct and weighted methods yielded less change than the stepwise methods in interpretation of the RDA results. Given the criticism of approaches relying on CWM traits we also applied fourth corner analysis with the original, harmonised, and aggregated trait datasets to detect relationships between traits and conductivity (methods and results in the supporting information \ref{sec:effects_harmonisation_trait_env_rel}).

@@

The calculation of FD metrics for the original, harmonised (not aggregated), and aggregated datasets suggested that harmonising trait data preserves FRic and FDiv values, whereas aggregated trait data only preserved FRic values. The FRic and FDiv values calculated from harmonised trait data correlated highly to FRic and FDiv values from the original dataset ($R^2 = 0.96$ and $0.86$, respectively). FRic and FDiv calculated for aggregated trait data showed moderate to high correlations to the FRic and FDiv values calculated with the original dataset (FRic: $0.78 <= R^2 <= 0.84$ , FDiv: $0.42 <= R^2 <= 0.62$, Figure \ref{fig:FRic} and \ref{fig:FDiv}). By contrast, FEve values calculated with harmonised and aggregated trait data were unrelated to the FEve values calculated with the original dataset (all $R^2 < 0.1$), except for \textit{stepwise\_agg \textsubscript{median}}, which showed a moderate correlation to FEve values calculated from the original dataset ($R^2  = 0.48$, Figure \ref{fig:FEve}).

@@

FD analysis quantifies the variability of traits that influence ecosystem functions. We found that harmonised trait data was well-suited for calculating FRic and FDiv values, and that aggregated trait data was well-suited for calculating FRic values. By contrast, the use of harmonised and aggregated trait data yielded completely different FEve values than with the trait data from \citet{szocs_effects_2014}. FRic represents the trait space spanned by the community (also termed the convex hull volume) and is only dependent on the trait data used. Our results show that harmonisation preserved the trait space and that aggregation only slightly changed the trait space. FDiv, which describes how abundances are distributed within the volume of the trait space, was preserved by harmonised trait data and to a lesser extent by aggregated trait data. FEve describes how regular the trait space is filled by taxa. We found that harmonised and aggregated trait changed the evenness of the abundance distribution in the trait space, which is likely because during the calculation of FEve only the nearest neighbours in the trait space are considered. This leads to the undesirable property that even small community changes result in large changes in FEve (\cite{botta-dukatTestingAbilityFunctional2016}).

@@

Although comprehensive freshwater invertebrate trait databases have been developed for several regions in the past, data synthesis presents challenges because of discrepancies in trait definitions, coding of trait affinities, and taxonomic resolution. By providing an overview of discrepancies in definitions, we have set a starting point for the development of standardised trait definitions through which invertebrate trait databases can be linked. Agreement on standard trait definitions and the subsequent development of ontologies are the next steps to facilitate trait-based analysis at large geographical scales. As our analysis showed, some grouping features might need to be re-classified to fit into such a controlled vocabulary. Although we showed that trait affinities from fuzzy and binary coding can be used together, we also recommend that a uniform coding of traits should be addressed during trait standardisation.

@@

\section*{Author contribution}
R.B.S. and S.K. designed the study; B.J.K. and L.M. provided trait data for Australia; A.S-K., W.G. and P.U-P. provided trait data for Europe; P.U-P. provided body form trait data for European and North American invertebrates; C.P.H., N.L.P., and L.T. provided trait data for North America; C.D.M. provided trait data for New Zealand; B.J.K., A.S-K., C.D.M, W.G., N.L.P., L.M., and C.P.H. provided in depth information on trait data and discussed the procedure of analysis; S.K. conducted data processing and the statistical analyses.; S.K. wrote the manuscript; B.J.K, A-S.K., C.D.M, P.U-P., C.P.H., and R.B.S. discussed the results and provided suggestions on the manuscript.
